Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-09 kappa.tex
Final worhty check before first milestone.

As ice flows and glides on its bed, it transports rock debris and erode the
landscape, thence leaving geomorphological traces of its former presence
(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
and early explorers have since long learned to read these traces, and to
understand that glaciers had once been more extensive than today. With the
advent of the glacial theory, birthed the idea that, under colder temperatures
than today, expansive \emph{ice sheets} had once covered much of Europe and
North America \citep{Agassiz.1840}.

However, the glacial theory did not gain much belief until the exploration and
discovery of the two present-day ice sheets on Earth, the Greenland and
Antarctic ice sheets, provided a modern analogue for the proposed European and
North American ice sheets. The first traverse of the Greenland was performed by
Nansen in 1888, while Amundsen reached the south pole in 1911. Meanwhile, in
the European Alps, it was then discovered from observations of the glacial
landscape, that there had not been only one, but at least four, glacial periods
\citep{Penck.Bruckner.1909}.

More recently, palaeo-climate records extracted from deep-sea sediments have
provided a much more detailed picture of the Earth environmental history
\citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973}, thus indicating neither
one, nor four, but rather tens of glacial cycles. During the last million
years, glacial and interglacial cycles have succeeded each other with a
100\,kyr periodicity \citep{Hays.etal.1976}. In the present thesis, we focus on
the most recent of these cycles: \emph{the last glacial cycle}
(\cref{fig:paleo-timeseries})

% ice volume 26.07 + 2.93 = 29.00
% meters sea-level 58.3 + 0.70 = 59.0
Together, the Greenland and Antarctic ice sheets currently host most glacial
ice on the planet, corresponding to 59~meters of potential sea-level rise
\citep{Bamber.etal.2001, Fretwell.etal.2013}. At the apogee of the last glacial
cycle, the last glacial maximum, sea-level was 120 to 135\,m lower than today
\citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
ice sheets and glaciers on land. Systematic mapping and compilation of glacial
landforms from all continents permitted reconstructions of this former ice
cover \citep{Ehlers.Gibbard.2007}, indicating that ice volume
was repartitioned between more extensive Greenland and Antarctic ice sheets,
the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
Innuitian, Laurentide and \emph{Cordilleran} ice sheets over North America
(\cref{fig:paleo-glaciation}).

Nevertheless, the landscape record is sparse in time and most often spatially
incomplete. Palaeo-ice sheets did not leave a systematic record of their
passage, and much of the evidence has been overridden by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can help.
Although these models alone contain too many uncertainties to reproduce
accurate picture of past glacial cycles \citep[e.g.,][]{Hebeler.etal.2008},
they can be calibrated against geological data where they exist, while
allowing a physics-based, quantitative reconstruction where they do not.

@@

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
(\cref{fig:map-northamerica}). The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
the Western Cordillera, which forms our region of focus in the present thesis.

@@

The North American Cordillera forms a broad mountain system with an
average east-west width of about 1000\,km. Its topography is a complex
jigsaw of mountain ranges and interior basins (\cref{fig:map-cordillera}).
It was formed during multiple accretion phases associated with the closure of
several back-arc oceanic trenches over a 200\,Ma-long geological history
\citep{Sigloch.Mihalynuk.2013}.

Most of the dominant reliefs are found along two major ribbons that run
parallel to each other along the general orientation of the Cordillera
(\cref{fig:map-cordillera}). The western ribbon is formed by the Pacific Coast
orogenic belt. Major glaciated reliefs include,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench \mref. It includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
Brooks Range, in northern Alaska, has been covered by ice, but its glaciers
were not connected to the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

@@

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
of the irregular arrangement of valleys and ridges that formed the ice sheet's
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them
\citep{Davis.Mathews.1944, Kleman.etal.2010}. In terms of numerical glacier
modelling, this topography implies complex patterns of flow and stress
within the ice, and requires high-resolution, and thus, large computational
grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effect is often
clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
large latitudinal extent of the region, mean annual temperatures vary
significantly from north to south (\cref{fig:plot-atm}a). About half of the
model domain has mean annual temperature below freezing. However, seasonality,
here computed as the temperature difference between the warmest and the coldest
months, varies between around 10\,\degC\ along most of the Pacific Coast to
above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperature is well below
freezing point do experience warm summers.

Precipitation rates are highest on the western slopes of the Pacific Coast
ranges and decline quickly on the other side of the mountains
(\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
even reinforced by the difference in timing of the precipitation peak through
the year (\cref{fig:plot-atm}d). Coastal regions receive most precipitation
during winter and fall, thus as snow. On the contrary, many inland regions
experience dry winters, and most of the annual precipitation, there,
consequently falls
as rain during the ablation season. The climatic setting of the North American
Cordillera is described further in \CCLI, where its effects on the ice sheet
model response have been studied in detail.

@@

In the northern sector of the ice sheet, at least four glaciations have been
identified
    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
           Briner.Kaufman.2008, Demuro.etal.2012,
           Stroeven.etal.2010, Stroeven.etal.2014}.
The two older ones, often collectively referred to as the Reid glaciations,
pre-date the last glacial cycle, and thus are beyond the scope of this study.
The two younger glaciations, though, occurred both during the last glacial
cycle. First, the Gladstone glaciation reached its maximum between
$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
during a period of high global ice volume on Earth, known as the Marine Isotope
Stage (MIS)~4. However, evidence from the Gladstone glaciation has not been
found in the southern part of the ice-covered region, making it difficult to
seize its magnitude. Second, the McConnell glaciation is associated with end
moraines of oldest minimum apparent exposure ages of
$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
which corresponds to the last global ice volume maximum (MIS~2). In contrary to
the Gladstone glaciation, the McConnell glaciation has been correlated with a
contemporaneous advance of the southern ice sheet margin, known as the Fraser
glaciation, with a maximum extent framed between
$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep[Fig.~4]{Porter.Swanson.1998}.
Within errors inherent to the dating methods \citep{Heyman.etal.2011}, the
McConnell and Fraser glaciations correspond to the same glacial episode, which
resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet.
This is the period where most geological observations point at.

@@

The southern margin of the ice sheet has been studied extensively, and is in
several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of different piedmont
lobes that have formed at the outlet of the main valleys. The Puget lobe
\citep[e.g.]{Thorson.1980, Porter.Swanson.1998} was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape, known as the Channeled Scabland.

The western margin of the ice sheet covered part of the continental shelf. An
ice cap covering Vancouver Island was well-connected to main body of ice on the
continent \mref. A significant part of the ice sheet thus drained into the Georgia
Strait and engulfed between Vancouver Island and the Olympic Peninsula where
it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
Further to the North, the ice margin reached Queen Charlotte Islands, yet the
archipelago may not have been fully covered by ice
flowing from the land. Submarine troughs likely indicate the erosional action of
former ice streams, yet it is not clear whether they have formed during the
last glacial cycle or earlier \mref.

Finally, the eastern margin of the ice sheet is the most unclear. Although it
is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
the nature of the interaction between the two ice sheets and the precise
location of this junction is unclear, while little information exists on its
timing
    \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}.
Because the
present thesis focuses specifically on the Cordilleran ice sheet, our largest
assumption certainly consists of neglecting buttressing effects from the
Laurentide ice sheet.

The ice sheet is believed to have retreated quickly. Most of the evidence comes
from the Fraser Plateau, where it indicates that some of the highest relief
emerged from the ice prior to the lowlands, possibly resulting in thin,
stagnant ice remnants in the interior
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
Much less is known about ice sheet decay in the northern parts of the
ice-covered areas, but glacial landforms from the Liard Lowland suggest a
different mode of deglaciation, with active retreat of the ice lobe towards the
surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in \CCYC.

@@

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
density, $\rho$, is constant, and the
conservation of mass implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$ \cref{fig:model-variables}, by
\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0 \,.
\end{equation}

@@

Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and $B=A^{-1/n}$, the
constitutive law (\ref{eqn:glenslaw}) can be rewritten using a sometimes more
convenient, inverted formulation,
\begin{equation}
    \label{eqn:invglenslaw}
    \DST = B\,\doteps_e^{1/n-1}\,\SRT \,.
\end{equation}
By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
by $\DST = 2 \nu\,\SRT$, thus yielding

@@

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation \citep[Eqs.~20--21]{Aschwanden.etal.2012} in order to
account for thermodynamic effects associated with internal phase changes in the
glacier. This enthalpy ($H$) formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the enthalpy can be
expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation
(\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,,
\end{equation}
where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
conductivity. Constant parameters of the ice flow model use in this thesis are
given in \cref{tab:iceflowparams}.

@@

PISM, the ice sheet model that we use, employs two well-documented
approximations of ice flow equations: the shallow ice approximation
    \citep[SIA,][]{Morland.Johnson.1980, Hutter.1983}
and the the shallow shelf approximation
    \citep[SSA,][]{Morland.1987, Weis.etal.1999}.
Although both approximations can be derived
from a rigorous scaling approach, we don't expand these lengthy derivations
here, assuming instead the simplifications and focusing on their effect.

@@

\todo{Add a few words about the fact that $\vsia$ can be directly obtained
      after one vertical integration. Show that vertical velocities can be
      obtained from the compressibility equation.}

@@

In PISM, the SIA and SSA velocities are eventually added
\citep[Eq.~15]{Winkelmann.etal.2011},
\begin{equation}
    \label{eqn:siassa}
    \vv = \vsia + \vssa \,.
\end{equation}
Because $\vsia$ is small where $\vssa$ is large and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met
\cref{fig:model-siassa}. However,
\cref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow and longitudinal flow has not yet been
mathematically proven

@@

The simulations presented in this thesis use a pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}} \,,
\end{equation}
where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law
\citep[Eq.~11]{Winkelmann.etal.2011},
\begin{equation}
    \label{eqn:plastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|} \,.
\end{equation}

@@

The modelled ice sheet interacts with the underlying bedrock in two ways.
Firstly, bedrock temperatures are computed to a given depth by appending the
ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
model. The only process accounted for is heat conduction, hence this model
consists of a simple diffusion equation,
\begin{equation}
    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T \,,
\end{equation}
where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
specific heat capacity of the bedrock. Bedrock temperature is conditioned at
depth by a fixed upwards geothermal heat flux $q_G$. The bedrock thermal model
is necessary for modelling the Cordilleran ice sheet on multi-millennial
time-scales, because temperature fluctuations caused by climate change and
isolating effects of the ice sheet penetrate several hundred meters into the
rock, and eventually affect basal ice temperatures.

Secondly, bedrock elevation evolve in response to the ice load. The bedrock
deformation model used in this thesis describes the flexure of an elastic
lithosphere on top of an infinite half-space viscous mantle
\citep{Lingle.Clark.1985}. It can be described by a single differential
equation of the bed elevation, $b$,
\begin{equation}
    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
\end{equation}
where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic (Laplacian
square) operator of
linear elastic theory, while $|\grad|$ is a pseudo-differential operator
defined through the Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. In
the left-hand part of this equation, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response. Parameter values for
bedrock response model are given in \cref{tab:bedrockparams}.

@@

\todo{Figure. Boundary models data flow chart}

>>>>

As ice flows and glides on its bed, it transports rock debris and erodes the
landscape, thence leaving geomorphological traces of its former presence
(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
and early explorers have since long learned to read these traces, and to
understand that glaciers had once been more extensive than today. With the
advent of the glacial theory, birthed the idea that, under colder temperatures
than today, expansive \emph{ice sheets} had once covered much of Europe and
North America \citep{Agassiz.1840}.

However, the glacial theory did not gain much belief until the exploration and
discovery of the two present-day ice sheets on Earth: the Greenland and
Antarctic ice sheets, provided a modern analogue for the proposed European and
North American ice sheets. The first traverse of Greenland was performed by
Nansen in 1888, while Amundsen reached the south pole in 1911. Meanwhile, in
the European Alps, it was then discovered from observations of the glacial
landscape, that there had not been only one, but at least four, glacial periods
\citep{Penck.Bruckner.1909}.

More recently, palaeo-climate records extracted from deep-sea sediments have
provided a much more detailed picture of the Earth environmental history
\citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973}, thus indicating neither
one, nor four, but rather tens of glacial cycles. During the last million
years, glacial and interglacial cycles have succeeded each other with a
100\,kyr periodicity \citep{Hays.etal.1976}. In the present thesis, we focus on
the most recent of these periods: \emph{the last glacial cycle}
(\cref{fig:paleo-timeseries})

% ice volume 26.07 + 2.93 = 29.00
% meters sea-level 58.3 + 0.70 = 59.0
Together, the Greenland and Antarctic ice sheets currently host most glacial
ice on the planet, corresponding to 59~meters of potential sea-level rise
\citep{Bamber.etal.2001, Fretwell.etal.2013}. At the apogee of the last glacial
cycle, the \emph{Last Glacial Maximum} (LGM), sea-level was 120 to 135\,m lower
than today
\citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
ice sheets and glaciers on land. Systematic mapping and compilation of glacial
landforms from all continents permitted reconstructions of this former ice
cover \citep{Ehlers.Gibbard.2007}, indicating that ice volume
was repartitioned between more extensive Greenland and Antarctic ice sheets,
the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
Innuitian and Laurentide ice sheets and, finally, the \emph{Cordilleran ice
sheet}, over North America (\cref{fig:paleo-glaciation}).

Nevertheless, the landscape record is sparse in time and most often spatially
incomplete. Palaeo-ice sheets did not leave a systematic record of their
passage, and much of the evidence has been overridden by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can help.
Although these models alone contain too many uncertainties to reproduce
accurate picture of past glacial cycles \citep[e.g.,][]{Hebeler.etal.2008},
they can be calibrated against geological data where they exist, while
allowing a physics-based, quantitative reconstruction where they do not
\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.

@@

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
(\cref{fig:map-northamerica}). The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, \emph{the Cordilleran ice sheet} capped the mountain
ranges of the Western Cordillera, which forms our region of focus in the
present thesis.

@@

The North American Cordillera forms a broad mountain system with an
average east-west width of about 1000\,km. Its topography is a complex
jigsaw of mountain ranges and interior basins (\cref{fig:map-cordillera}).
It was formed during multiple accretion phases associated with the closure of
several back-arc oceanic trenches, over a 200\,Ma-long geological history
\citep{Sigloch.Mihalynuk.2013}.

Most of the dominant reliefs are found along two major ribbons that run
parallel to each other along the general orientation of the Cordillera
(\cref{fig:map-cordillera}). The western ribbon is formed by the Pacific Coast
orogenic belt \mref. Major glaciated reliefs include,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench \mref. It includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
Brooks Range, in northern Alaska, has been covered by ice, but its glaciers
were not connected to the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

@@

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
of the irregular arrangement of valleys and ridges that formed the ice sheet's
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them
\citep{Davis.Mathews.1944, Kleman.etal.2010}. In terms of numerical glacier
modelling, this topography implies complex patterns of flow and stress
within the ice and requires high-resolution, and thus, large computational
grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effect is often
clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
large latitudinal extent of the region, \emph{mean annual temperatures} vary
significantly from north to south (\cref{fig:plot-atm}a). About half of the
model domain has mean annual temperature below freezing. However,
\emph{seasonality},
here computed as the temperature difference between the warmest and the coldest
months, varies between around 10\,\degC\ along most of the Pacific Coast to
above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperature is well below
freezing point do experience warm summers.

Rates of \emph{precipitation} are highest on the western slopes of the Pacific
Coast
ranges and decline quickly on the eastern side of the mountains
(\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
even reinforced by the difference in timing of the \emph{precipitation peak}
through
the year (\cref{fig:plot-atm}d). Coastal regions receive most precipitation
during winter and fall, thus as snow. On the contrary, many inland regions
experience dry winters, and most of the annual precipitation,
consequently falls there
as rain during the ablation season. The climatic setting of the North American
Cordillera is described further in \CCLI, where its effects on the ice sheet
model response have been studied in detail.

@@

In the \emph{northern sector} of the ice sheet, at least four glaciations have
been identified
    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
           Briner.Kaufman.2008, Demuro.etal.2012,
           Stroeven.etal.2010, Stroeven.etal.2014}.
The two older ones, often collectively referred to as the Reid glaciations,
pre-date the last glacial cycle, and thus are beyond the scope of this study.
The two younger glaciations, though, occurred both during the last glacial
cycle. First, the Gladstone glaciation reached its maximum between
$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
during a period of high global ice volume on Earth, known as the Marine Isotope
Stage (MIS)~4. However, evidence from the Gladstone glaciation has not been
found in the southern part of the ice-covered region, making it difficult to
seize its magnitude. Second, the McConnell glaciation is associated with end
moraines of oldest minimum apparent exposure ages of
$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
which corresponds to the last global ice volume maximum (MIS~2). In contrary to
the Gladstone glaciation, the McConnell glaciation has been correlated with a
contemporaneous advance of the southern ice sheet margin, known as the Fraser
glaciation, with a maximum extent framed between
$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep[Fig.~4]{Porter.Swanson.1998}.
Within errors inherent to the dating methods \citep{Heyman.etal.2011}, the
McConnell and Fraser glaciations correspond to the same glacial episode, which
resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet.
This is the period where most geological observations point at.

@@

The \emph{southern margin} of the ice sheet has been studied extensively, and
is in several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of different piedmont
lobes that have formed at the outlet of the main valleys. The Puget lobe
\citep[e.g.]{Thorson.1980, Porter.Swanson.1998} was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape, known as the Channeled Scabland.

The \emph{western margin} of the ice sheet covered part of the continental
shelf. An
ice cap covering Vancouver Island was well-connected to main body of ice on the
continent \mref. A significant part of the ice sheet thus drained into the Georgia
Strait and engulfed between Vancouver Island and the Olympic Peninsula where
it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
Further to the North, the ice margin reached Queen Charlotte Islands, yet the
archipelago may not have been fully covered by ice
flowing from the land. Submarine troughs likely indicate the erosional action of
former ice streams, yet it is not clear whether they have formed during the
last glacial cycle or earlier \mref.

Finally, the \emph{eastern margin} of the ice sheet is the most unclear.
Although it
is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
the nature of the interaction between the two ice sheets and the precise
location of this junction is unclear, while little information exists on its
timing
    \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}.
Because the
present thesis focuses specifically on the Cordilleran ice sheet, our largest
assumption certainly consists of neglecting buttressing effects from the
Laurentide ice sheet.

The ice sheet is believed to have retreated quickly. Most of the evidence comes
from the Fraser Plateau, where it indicates that some of the highest relief
emerged from the ice prior to the lowlands, possibly resulting in thin,
stagnant ice remnants in the interior
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
Much less is known about ice sheet decay in the northern parts of the
ice-covered areas, but glacial landforms from the Liard Lowland suggest a
different mode of \emph{deglaciation}, with active retreat of the ice lobe
towards the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in \CCYC.

@@

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
density, $\rho$, is constant, and the
conservation of mass implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0 \,.
\end{equation}

@@

Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and $B=A^{-1/n}$, the
constitutive law (\ref{eqn:glenslaw}) can be rewritten using a sometimes more
convenient, inverted formulation,
\begin{equation}
    \label{eqn:invglenslaw}
    \DST = B\,\doteps_e^{1/n-1}\,\SRT \,.
\end{equation}
By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
by $\DST = 2 \nu\SRT$, thus yielding

@@

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation \citep[Eqs.~20--21]{Aschwanden.etal.2012} in order to
account for thermodynamic effects associated with internal phase changes in the
glacier. This enthalpy ($H$) formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the enthalpy can be
expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation
(\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,,
\end{equation}
where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
conductivity. Parameter values for the ice flow model used in this thesis are
given in \cref{tab:iceflowparams}.

@@

PISM, the ice sheet model that we use, employs two well-documented
approximations of ice flow equations: the shallow ice approximation (SIA)
and the the shallow shelf approximation (SSA). Although both approximations can
be derived from a rigorous scaling approach
    \citep[SIA,][]{Morland.Johnson.1980, Hutter.1983,
                   Morland.1987, Weis.etal.1999},
we don't expand these lengthy derivations
here, assuming instead the simplifications and focusing on their effect.

@@

\todo{Add a few words to emphasize that $\vsia$ can be directly obtained
      after one vertical integration. Show that vertical velocities can be
      obtained from the compressibility equation.}

@@

\todo{Add a few words to emphasize that $\vssa$ can not be obtained in a
      direct manner, requiring to solve a linear system.}

In PISM, the SIA and SSA velocities are eventually added
\citep[Eq.~15]{Winkelmann.etal.2011},
\begin{equation}
    \label{eqn:siassa}
    \vv = \vsia + \vssa \,.
\end{equation}
Because $\vsia$ is small where $\vssa$ is large and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met
(\cref{fig:model-siassa}). However,
\cref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow and longitudinal flow has not been
mathematically proven.

@@

The simulations presented in this thesis use a \emph{pseudo-plastic sliding
law},
\begin{equation}
    \label{eqn:pseudoplastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}} \,,
\end{equation}
where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law
\citep[Eq.~11]{Winkelmann.etal.2011},
\begin{equation}
    \label{eqn:plastic}
    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|} \,.
\end{equation}

@@

The modelled ice sheet interacts with the underlying bedrock in two ways.
Firstly, \emph{bedrock temperatures} are computed to a given depth by
appending the
ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
model. The only process accounted for is heat conduction, hence this model
consists of a simple diffusion equation,
\begin{equation}
    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T \,,
\end{equation}
where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
specific heat capacity of the bedrock. Bedrock temperature is conditioned at
depth by a fixed upwards geothermal heat flux $q_G$. The bedrock thermal model
is necessary for modelling the Cordilleran ice sheet on multi-millennial
time-scales, because temperature fluctuations caused by climate change and
isolating effects of the ice sheet penetrate several hundred meters into the
rock, and eventually affect basal ice temperatures.

Secondly, bedrock elevation evolve in response to the ice load. The
\emph{bedrock deformation} model used in this thesis describes the flexure of
an elastic
lithosphere on top of an infinite half-space viscous mantle
\citep{Lingle.Clark.1985}. It can be described by a single pseudo-differential
equation of the bed elevation, $b$,
\begin{equation}
    \label{eqn:beddef}
    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
\end{equation}
where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
\citep{Bueler.etal.2007}. Here, ${\Delta^2 = (\div\grad{})^2}$ designs the
biharmonic (Laplacian square) operator of
linear elastic theory, while $|\grad{}|$ is a pseudo-differential operator
defined through its Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. On
the left-hand side of \cref{eqn:beddef}, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response. Parameter values for
bedrock response model are given in \cref{tab:bedrockparams}.

@@
